---
title: "Hands-on Exercise 7b - Visualising Geospatial Point Data"
date: February 25, 2024
author: "Imran Ibrahim"
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# 7.1 Overview

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people. Like choropleth maps, you can create classed or unclassed versions of these maps. The classed ones are known as range-graded or graduated symbols, and the unclassed are called proportional symbols, where the area of the symbols are proportional to the values of the attribute being mapped. In this hands-on exercise, we will learn how to create a proportional symbol map showing the number of wins by Singapore Pools’ outlets using an R package called **tmap**.

## 7.1.1 Learning outcome

By the end of this hands-on exercise, we will acquire the following skills by using appropriate R packages:

-   To import an aspatial data file into R.

-   To convert it into simple point feature data frame and at the same time, to assign an appropriate projection reference to the newly create simple point feature data frame.

-   To plot interactive proportional symbol maps.

# 7.2 Loading R packages and Data preparation

Before we get started, we need to ensure that **tmap** package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 7.2.1 The data

The data set used is called *SGPools_svy21*.csv

It consists of seven columns. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches. They are in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

## 7.2.2 Data import and preparation

First, we will import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
```

Next, we examine if the data file has been imported correctly.

```{r}
list(sgpools) 
```

## 7.2.3 Creating a sf data frame from an aspatial data frame

The code below converts sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** packages

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

Things to learn from the arguments above:

-   The *coords* argument requires us to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.

-   The *crs* argument required us to provide the coordinates system in epsg format. [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System. You can search for other country’s epsg code by refering to [epsg.io](https://epsg.io/).

We can display the basic information of the newly created sgpools_sf below.

```{r}
list(sgpools_sf)
```

The output shows that sgppols_sf is in point feature class. It’s epsg ID is 3414. The bbox provides information of the extend of the geospatial data.

# 7.3 Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

## 7.3.1 Starting with an interactive point symbol map

The code below is used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = 1,
           border.col = "black",
           border.lwd = 1)
```

## 7.3.2 Making it proportional

To draw a proportional symbol map, we need to assign a numerical variable to the size visual attribute. The code below show that the variable *Gp1Gp2Winnings* is assigned to size visual attribute.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = "Gp1Gp2 Winnings",
           border.col = "black",
           border.lwd = 1)
```

## 7.3.3 Using a different colour 

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunks below, *OUTLET_TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)
```

## 7.3.4 Faceted plots using  `tm_facets()` method

The argument *sync* in *tm_facets()* can be used to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Finally, we will switch **tmap**’s Viewer back to plot mode by using the code below.

```{r}
tmap_mode("plot")
```

## 7.3.5 Plotting Practise

Below are some additional plots for practise.

# 7.4 References

Main reference: Kam, T.S. (2024). [Visualising Geospatial Point Data.](https://r4va.netlify.app/chap22#overview)
